Physcraper: Body

### 2020-07-09 2-body.Rmd
intro re-structure

More than a decade ago, the National Center for Biodiversity Information (NCBI) molecular database, GenBank, released its version number 159 (April 15, 2007). With 72 million DNA sequences, it was estimated to have the potential to resolve evolutionary relationships of most of the 241 000 distinct taxa represented in it [about `r round(236023/240708, digits=4)*100`% of taxa in the NCBI taxonomy; @sanderson2008phylota], which in turn covers about 10% of extant described biodiversity [taking a conservative estimate of extant diversity; @scott2011ncbi; @federhen2003taxonomy].
In comparison, the current GenBank release number 238 (June 15, 2020) has tripled in size, hosting data for more than 217 million DNA sequences [(See GenBank's release website)](https://ftp.ncbi.nih.gov/genbank/gbrel.txt). Yet, we only have data on phylogenetic relationships available for about 90 000 taxa [@hinchliff2015synthesis; CURRENT SYNTH TREE CITATION], covering less than one third of the taxonomic diversity with data available a decade ago. Some of this discrepancy is explained by the many phylogenies that are not publicly shared [@drew2013lost; @magee2014dawn, @mctavish2018bioessay]. However, most recently published large trees have been made available in recent years, indicating there is also a lag between the amount of new DNA data generated and the analysis of these data in a phylogenetic context.

@@

- In order to take advantage of the OpenTree tools, the input tree needs to be stored in the
  OpenTree phylesystem [github.com/opentreeoflife/phylesystem](https://github.com/opentreeoflife/phylesystem).
  The main
  reason for this is that trees in phylesystem have a set of user defined characteristics
  that are essential for automatizing the phylogeny update process. The most relevant of these being the standardization of taxonomic names and the definition of ingroup and outgroup. Outgroup and ingroup taxa in the original tree are identified and tagged. This allows to automatically set the root for the updated tree on the final steps of the pipeline.
  Currently, only trees connected to
  a published study can be stored in the phylesystem (although there are plans to
  allow storage of unpublished trees).
  A user can choose from among the `r tryCatch({rotl::tol_about()$num_source_trees}, error = function(e) {1216})` published trees supporting the resolved nodes of the synthetic tree in the OpenTree website ([See OpenTree's website about](https://tree.opentreeoflife.org/about/synthesis-release/v12.3)). If the published tree you are interested in updating is not in there, you can upload it via OpenTree's curator tool ([www.opentreeoflife.org/curator](https://tree.opentreeoflife.org/curator).
- THE INPUT TREE CAN ALSO BE A PERSONAL ONE. A TNRS MATCHING NEEDS TO BE PERFORMED (AND A FILE GENERATED) IN ORDER TO UPDATE A PERSONAL UNPUBLISHED TREE.
- The alignment should be a gene alignment that was used to generate the tree. The original
  alignments are usually stored in a public repository such as TreeBase [@piel2009treebase; @vos2012nexml],
  DRYAD ([www.datadryad.org](http://datadryad.org/)), or the journal were the tree was originally published.
  If the alignment is stored in TreeBase, Physcraper can download it directly,
  either from the TreeBASE website ([www.treebase.org](https://treebase.org/))
  or through the TreeBASE GitHub repository (SuperTreeBASE; [github.com/TreeBASE/supertreebase](https://github.com/TreeBASE/supertreebase)).
  If the alignment is on another repository, or constitutes personal data, a path to a local copy of the alignment has to be provided.
- A taxon name matching step is performed to verify that all taxon names on the tips
  of the tree are in the DNA character matrix and vice versa.
- A ".csv" file with the summary of taxon name matching is produced for the user.
- Unmatched taxon names are dropped from both the tree and alignment.
  Technically, just one matching name is needed to perform the searches. Please, see next section.
- A ".tre" file and a ".fas" file containing only the matched taxa are generated and saved in the `inputs` folder to be used in the following steps.

@@

Research was supported by the grant "Sustaining the Open Tree of Life", National Science Foundation ABI No. 1759838.
Computer time was provided by the Multi-Environment Research Computer for Exploration and Discovery (MERCED) cluster from the University of California, Merced (UCM), supported by the NSF Grant No. ACI-1429783.

>>>>

More than a decade ago, the National Center for Biodiversity Information (NCBI) molecular database, GenBank, released its version number 159 (April 15, 2007). With 72 million DNA sequences, it was estimated to have the potential to resolve evolutionary relationships of most of the 241 000 distinct taxa represented in it [about `r round(236023/240708, digits=4)*100`% of taxa in the NCBI taxonomy; @sanderson2008phylota], which in turns covers about 10% of extant described biodiversity [taking a conservative estimate of extant diversity; @scott2011ncbi; @federhen2003taxonomy].
In comparison, the current GenBank release number 238 (June 15, 2020) has tripled in size, hosting data for more than 217 million DNA sequences [(See GenBank's release website)](https://ftp.ncbi.nih.gov/genbank/gbrel.txt). Yet, publicly available estimates of phylogenetic relationships still cover only about 90 000 taxa [@hinchliff2015synthesis; CURRENT SYNTH TREE CITATION], covering less than one third of the taxonomic diversity with data available a decade ago. Some of this discrepance is because that many phylogenies are not publicly shared [@drew2013lost; @magee2014dawn, @mctavish2018bioessay]. However, most recently published large trees have been made available in recent years, indicating there is also a lag between the amount of new DNA data generated and the analysis of these data in a phylogenetic context.

@@

- In order to take advantage of the OpenTree tools, the input tree needs to be stored in the
  OpenTree phylesystem [github.com/opentreeoflife/phylesystem](https://github.com/opentreeoflife/phylesystem), or need to have the tip labels mapped to the unified OpenTree taxonomy via bulk taxonomic name resolution (https://tree.opentreeoflife.org/curator/tnrs/).
  The main
  reason for this is that trees in phylesystem have a set of user defined characteristics
  that are essential for automatizing the phylogeny update process. The most relevant of these being the standardization of taxonomic names and the definition of ingroup and outgroup. Outgroup and ingroup taxa in the original tree are identified and tagged. This allows to automatically set the root for the updated tree on the final steps of the pipeline.
  Currently, only trees connected to
  a published study can be stored in the phylesystem (although there are plans to
  allow storage of unpublished trees).
  A user can choose from among the `r tryCatch({rotl::tol_about()$num_source_trees}, error = function(e) {1216})` published trees supporting the resolved nodes of the synthetic tree in the OpenTree website ([See OpenTree's website about](https://tree.opentreeoflife.org/about/synthesis-release/v12.3)). If the published tree you are interested in updating is not in there, you can upload it via OpenTree's curator tool ([www.opentreeoflife.org/curator](https://tree.opentreeoflife.org/curator).
  If the user is not ready to make the input tree public, `physcraper` can use as input a tree, and aliognment, and a file of mappings between the tip labels and taxonomic names.
- The input alignment should be a single locus alignment that was used to generate the tree. For public data, the original
  alignments are often stored in a public repository such as TreeBase [@piel2009treebase; @vos2012nexml],
  DRYAD ([www.datadryad.org](http://datadryad.org/)), or the journal were the tree was originally published.
  If the alignment is stored in TreeBase, Physcraper can download it directly,
  either from the TreeBASE website ([www.treebase.org](https://treebase.org/))
  or through the TreeBASE GitHub repository (SuperTreeBASE; [github.com/TreeBASE/supertreebase](https://github.com/TreeBASE/supertreebase)).
  If the alignment is on another repository, or constitutes personal data, a path to a local copy of the alignment has to be provided.
- A taxon name matching step is performed to verify that all taxon names on the tips
  of the tree are in the DNA character matrix and vice versa.
- A ".csv" file with the summary of taxon name matching is produced for the user.
- Unmatched taxon names are dropped from both the tree and alignment.
  Technically, just one matching name is needed to perform the searches. Please, see next section.
- A ".tre" file and a ".fas" file containing only the matched taxa are generated and saved in the `inputs` folder to be used in the following steps.

@@

Research was supported by the grant "Sustaining the Open Tree of Life", National Science Foundation ABI No. 1759838, and ABI No. 1759846.
Compute time was provided by the Multi-Environment Research Computer for Exploration and Discovery (MERCED) cluster from the University of California, Merced (UCM), supported by the NSF Grant No. ACI-1429783.
